---
title: "Incremental test analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Incremental test analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)
```

If you have an incremental test, `whippr` can also help you out! In this vignette you will learn how to:

* Normalize your data (i.e., recognize baseline period, adjust work rates, and adjust time column);
* Recognize "bad breaths" and remove them;
* Calculate mean response time through various methods (*work in progress*);
* Calculate maximal oxygen uptake (and analyze whether a plateau existed in your data);
* Estimate ventilatory thresholds (*work in progress*).

All of the above work both in a **ramp** and in a **step** incremental test.

## Normalize your data

To get started, let's read in the example data from a ramp incremental test that comes with the package. This is a test that had the following protocol:

- 4-min baseline period at 20 W;
- ramp increase of 25 W/min.

```{r setup}
library(whippr)
library(ggplot2)
library(dplyr)

path_example <- system.file("ramp_cosmed.xlsx", package = "whippr")

df <- read_data(path = path_example, metabolic_cart = "cosmed")

df
```

```{r}
df %>%
  ggplot(aes(t, VO2)) +
  geom_point(shape = 21, size = 4, fill = "white") +
  theme_whippr()
```

A few things to note:

- Our time column is not normalized. That is, we do not know explicitly what is baseline, and what is ramp;
- We do not have a work rate column.

All of the above will be fixed with the `incremental_normalize()` function. Since we do not have a work rate column, we will set the argument `work_rate_magic = TRUE`. This argument will allow us to calculate the work rates throughout the test.

```{r}
ramp_normalized <- df %>%
 incremental_normalize(
   .data = .,
   incremental_type = "ramp",
   has_baseline = TRUE,
   baseline_length = 240, ## 4-min baseline
   work_rate_magic = TRUE,
   baseline_intensity = 20, ## baseline was performed at 20 W
   ramp_increase = 25 ## 25 W/min ramp
 )

ramp_normalized
```

We can see that now our data is aware of the two different phases in the test: baseline and ramp period:

```{r}
ramp_normalized %>% 
  distinct(protocol_phase)
```

Additionally, a new column was created: `work_rate`:

```{r}
ramp_normalized %>% 
  select(t, work_rate)
```

And we can also plot it to check what was done. Note that there is a constant-load during baseline (20 W), and then a constant increase in power output (25 W/min) during the ramp phase:

```{r}
ramp_normalized %>% 
  plot_incremental()
```

We can therefore quickly analyze our peak power output as:

```{r}
ramp_normalized %>% 
  slice_max(work_rate) %>% 
  select(work_rate)
```

### Step test

But what if we had a step-incremental test? How does the work rate transformation work? In a step test is important to have both the actual power output from each step, and also a linearization of the power output. To illustrate what I mean, let's take a look at an example:

```{r}
## get file path from example data
path_example_step <- system.file("step_cortex.xlsx", package = "whippr")

## read data from step test
df_step <- read_data(path = path_example_step, metabolic_cart = "cortex")

df_step
```

Note that our data contains weird column names. So, for simplicity, let's rename the VO2 column.

```{r}
df_step_renamed <- df_step %>% 
  rename(VO2 = `V'O2 (STPD)`)
```

That is a test that had the following protocol:

- resting period (i.e, 0 W) for 2 minutes;
- step protocol starting at 50 W and increasing 25 W every 3 minutes.

```{r}
df_step_renamed %>% 
  ggplot(aes(t, VO2)) +
  geom_point(shape = 21, size = 4, fill = "white") +
  theme_whippr()
```

So, let's first normalize our data:

```{r}
step_normalized <- df_step %>%
 incremental_normalize(
   .data = .,
   incremental_type = "step",
   has_baseline = TRUE,
   baseline_length = 120, ## 2 min baseline
   work_rate_magic = TRUE, 
   baseline_intensity = 0, ## baseline was resting on the bike, so intensity is 0 W
   step_start = 50, ## step protocol started at 50 W
   step_increase = 25, ## step increase was 25 W
   step_length = 180 ## the intensity increased every 3 minutes
 )

step_normalized
```

And then we can visualize what was done with the work rate with the `plot_incremental()` function:

```{r fig.width=10, fig.height=10}
step_normalized %>% 
  plot_incremental()
```

As you can note, two working rates were created: one with the actual power output, and another one with the linearization of the power output.

```{r}
step_normalized %>% 
  select(t, protocol_phase:step)
```

This is useful, for example, to calculate the peak power output:

```{r}
step_normalized %>% 
  slice_max(work_rate) %>% 
  select(work_rate)
```

## Recognize bad breaths

As you might have noticed, these two incremental test examples had a few bad breaths (outliers) that should be deleted prior to any data analysis. This can be easily achieved with the `detect_outliers()` function.

Two methods for detecting outliers are available: **linear** and **anomaly** detection. The **linear** method is going to fit two linear models: one for the baseline period, and another one for the ramp (or step) period. The **anomaly** detection, however, uses the [anomalize](https://business-science.github.io/anomalize/) package, which decompose time series, and then perform the anomaly detection. Let's see it in action:

### Linear

```{r}
## detect ramp outliers
data_ramp_outliers <- detect_outliers(
  .data = ramp_normalized,
  test_type = "incremental",
  vo2_column = "VO2",
  cleaning_level = 0.95,
  method_incremental = "linear",
  verbose = TRUE
 )

data_ramp_outliers
```

```{r}
data_ramp_outliers %>% 
  plot_outliers()
```

Thereafter, you can easily remove the detected outliers like this:

```{r}
data_ramp_outliers %>% 
  filter(outlier == "no")
```

Note that we set **95%** of confidence level when detecting the outliers. You can easily change that, for example, to **99%**:

```{r}
detect_outliers(
  .data = ramp_normalized,
  test_type = "incremental",
  vo2_column = "VO2",
  cleaning_level = 0.99, ## changed to 99%
  method_incremental = "linear",
  verbose = TRUE
 ) %>% 
  plot_outliers()
```

### Anomaly

Now let's see how the anomaly detection performs:

```{r}
detect_outliers(
  .data = ramp_normalized,
  test_type = "incremental",
  vo2_column = "VO2",
  cleaning_level = 0.95,
  method_incremental = "anomaly", ## changed to anomaly detection
  verbose = TRUE
 ) %>% 
  plot_outliers()
```

## Mean response time

> Work in progress.

## VO2max (maximal oxygen uptake)

There are two functions that you can use to analyze **VO2max**:

- `vo2_max()`: it performs all the necessary steps, which include:
    - `incremental_normalize()`: normalize incremental test data
    - `detect_outliers()`: detect outliers
    - `interpolate()`: interpolate data from breath-by-breath into second-by-second
    - `perform_average()`: perform average on second-by-second data
    
- `perform_max()`: it only performs the final steps (`interpolate()` and `perform_average()`).

### `perform_max()`

```{r}
results_vo2max <- data_ramp_outliers %>% ## data was already normalized and outliers were detected
  perform_max(
    .data = .,
    vo2_column = "VO2",
    vo2_relative_column = "VO2/Kg",
    heart_rate_column = "HR",
    rer_column = "R",
    average_method = "bin",
    average_length = 30,
    plot = TRUE,
    verbose = FALSE
  )

results_vo2max
```

```{r}
results_vo2max$plot[[1]]
```

### `vo2_max()`

```{r}
vo2_max(
  .data = df, ## data from `read_data()`
  vo2_column = "VO2",
  vo2_relative_column = "VO2/Kg",
  heart_rate_column = "HR",
  rer_column = "R",
  detect_outliers = TRUE,
  average_method = "bin",
  average_length = 30,
  plot = TRUE,
  verbose = TRUE,
  ## arguments for `incremental_normalize()`
  incremental_type = "ramp",
  has_baseline = TRUE,
  baseline_length = 240, ## 4-min baseline
  work_rate_magic = TRUE, ## produce a work rate column
  baseline_intensity = 20, ## baseline was performed at 20 W
  ramp_increase = 25, ## 25 W/min ramp
  ## arguments for `detect_outliers()`
  test_type = "incremental",
  cleaning_level = 0.95, 
  method_incremental = "linear"
)
```


## Ventilatory thresholds

> Work in progress.
